---
title: "Statistical Analysis With Moran's I"
output: html_document
date: "2026-02-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Global Moran's I for diabetes mortality (province level)

## 1. Loading libraries
```{r}
# Libraries 
library(sf)
library(dplyr)
library(spdep)
library(tmap)
library(viridis)
library(ggplot2)

```

## 2. Data
```{r}
# Load data 
provinces_full <- readRDS("../data_clean/provinces_full.rds")
# Basic checks
stopifnot(inherits(provinces_full, "sf"))
stopifnot("diabetes_mort_rate" %in% names(provinces_full))
```

## 3. Choose a projected CRS for spatial operations
```{r}
# Moran's I uses neighbors, so CRS is not as critical as distance-based methods,
# but projecting avoids potential geometry issues.
# ETRS89 / LAEA Europe (EPSG:3035) is a standard choice for EU-wide work.
prov_proj <- st_transform(provinces_full, 3035)

# Optional: fix invalid geometries if needed
if (any(!st_is_valid(prov_proj))) {
  prov_proj <- st_make_valid(prov_proj)
}
```

## 4. Build spatial weights (queen contiguity) 
```{r}
# Queen contiguity: provinces are neighbors if they share a border OR a point.
nb_q <- poly2nb(prov_proj, queen = TRUE)

# Check for "islands" (units with no neighbors)
no_nb <- which(card(nb_q) == 0)
if (length(no_nb) > 0) {
  message("Provinces with zero neighbors (islands): ",
          paste(prov_proj$prov_name[no_nb], collapse = ", "))
}

# Create weights list
lw_q <- nb2listw(nb_q, style = "W", zero.policy = TRUE)

```
## 5. Global Moran's I
```{r}

y <- prov_proj$diabetes_mort_rate

# Moran test (randomization)
moran_res <- moran.test(y, lw_q, zero.policy = TRUE)
print(moran_res)
```
It's strongly positive, so there is clustering.

It means that provinces with high diabetes mortality tend to be located next to other high-mortality provinces, and low-mortality provinces tend to cluster together.

## 6. Permutation test (more robust)
```{r}
set.seed(123)
moran_mc <- moran.mc(y, lw_q, nsim = 999, zero.policy = TRUE)
print(moran_mc)

```
Out of 999 random permutations, none produced a Moran’s I as large as the observed one. This confirms the result.

# 7) Save results 
```{r}
moran_out <- data.frame(
  statistic = "Global Moran's I",
  I = as.numeric(moran_res$estimate[["Moran I statistic"]]),
  expectation = as.numeric(moran_res$estimate[["Expectation"]]),
  variance = as.numeric(moran_res$estimate[["Variance"]]),
  p_value = moran_res$p.value,
  mc_p_value = moran_mc$p.value,
  nsim = 999
)

write.csv(moran_out, "outputs/moransI_diabetes.csv", row.names = FALSE)

```

# 8. Moran scatterplot 
```{r}
# This helps interpretation: slope ~ Moran's I
png("../outputs/moran_scatter_diabetes.png", width = 1200, height = 900, res = 140)
moran.plot(y, lw_q, zero.policy = TRUE,
           labels = prov_proj$prov_name,
           xlab = "Diabetes mortality rate",
           ylab = "Spatial lag of mortality rate",
           main = "Moran scatterplot: diabetes mortality")
dev.off()

```

# LISA (Local Moran's I)

```{r}
# 1 Local Moran statistics
lisa <- localmoran(y, lw_q, zero.policy = TRUE)

# localmoran() columns:
# Ii      = local Moran's I
# E.Ii    = expectation
# Var.Ii  = variance
# Z.Ii    = Z-score
# Pr(z != E(Ii)) = pseudo p-value (normal approximation)
prov_proj <- prov_proj %>%
  mutate(
    lisa_I   = lisa[, "Ii"],
    lisa_z   = lisa[, "Z.Ii"],
    lisa_p   = lisa[, "Pr(z != E(Ii))"]
  )
# 2. Compute spatial lag of y (needed for quadrant classification)
lag_y <- lag.listw(lw_q, y, zero.policy = TRUE)

# Standardize y and lag_y (z-scores)
y_z <- as.numeric(scale(y))
lag_y_z <- as.numeric(scale(lag_y))

prov_proj <- prov_proj %>%
  mutate(
    y_z = y_z,
    lag_y_z = lag_y_z
  )

```


```{r}

# 3. Cluster classification (HH, LL, HL, LH) with significance threshold
alpha <- 0.05

prov_proj <- prov_proj %>%
  mutate(
    lisa_cluster = case_when(
      lisa_p <= alpha & y_z >= 0 & lag_y_z >= 0 ~ "High-High",
      lisa_p <= alpha & y_z <= 0 & lag_y_z <= 0 ~ "Low-Low",
      lisa_p <= alpha & y_z >= 0 & lag_y_z <= 0 ~ "High-Low",
      lisa_p <= alpha & y_z <= 0 & lag_y_z >= 0 ~ "Low-High",
      TRUE ~ "Not significant"
    )
  )

# Optional: set factor order for nicer legends
prov_proj$lisa_cluster <- factor(
  prov_proj$lisa_cluster,
  levels = c("High-High", "Low-Low", "High-Low", "Low-High", "Not significant")
)

```

## 4. Quick diagnostic counts
```{r}
print(table(prov_proj$lisa_cluster))

```

## 5. Save LISA results table
```{r}

lisa_table <- prov_proj %>%
  st_drop_geometry() %>%
  select(
    prov_code, prov_name, region_name,
    diabetes_mort_rate,
    lisa_I, lisa_z, lisa_p,
    lisa_cluster
  ) %>%
  arrange(lisa_p)

write.csv(lisa_table, "outputs/lisa_diabetes_table.csv", row.names = FALSE)

```

```{r}
# 6. LISA cluster map (ggplot)
# Note: default ggplot colors are fine (no need to set manual colors)
prov_plot <- st_transform(prov_proj, 4326)

p_lisa <- ggplot(prov_plot) +
  geom_sf(aes(fill = lisa_cluster), color = "grey30", linewidth = 0.1) +
  labs(
    title = "LISA clusters for diabetes mortality (Local Moran's I)",
    subtitle = paste0("Queen contiguity weights; significance alpha = ", alpha),
    fill = "Cluster type",
    caption = "High-High/Low-Low indicate significant local clustering; High-Low/Low-High indicate spatial outliers."
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold")
  )

ggsave(
  filename = "outputs/lisa_clusters_diabetes.png",
  plot = p_lisa,
  width = 10,
  height = 7,
  dpi = 200
)

p_lisa

```

